---
title: "Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(parray)
```

## Introduction

The `parray` package provides a set of functions for manipulating arrays of probabilities that are 
dependent or conditioned on a set of discrete variables.


### mtcars

For the `mtcars` dataset, to calculate the probability of a particular combination of numbers of 
cylinders and carburetors for a given number of gears $P(\text{cyl},\text{carb}\,|\,\text{gear})$
```{r}
ptabs(~carb+cyl|gear,data=mtcars)
```
This produces a named array with dimensions `carb`, `cyl` and `gear`.  The marginal 
probabilities of the number of gears is
```{r}
ptabs(~gear,data=mtcars)
```
and the product of these conditional and marginal probabilities
```{r}
product(ptabs(~cyl+carb|gear,data=mtcars),ptabs(~gear,data=mtcars))
```
is the joint distribution of the number cylinders, carburetors and gears
```{r}
ptabs(~cyl+carb+gear,data=mtcars)
```

The `ptabs` function can take a dataframe or a named array as its `data` argument.  If `data` is a 
named array that is already conditioned on a set of variables, then those conditioning varaibles are 
retained in the result, but dependent variables that do not appear in the formula are marginalized
over 
```{r}
ptabs(~cyl,data=ptabs(~cyl+carb|gear,data=mtcars))
```

In the `mtcars` dataset, some combinations of cylinders and carburetors do not occur
```{r}
xtabs(~cyl+carb,data=mtcars)
```
The conditional probabilities for the missing combinations are undefined and `ptabs` replaces them
with the value of the `fill` argument.  By default `fill` is zero, but may be a vector or a named 
array with compatible dimensions
```{r}
ptabs(~gear|cyl+carb,data=mtcars,fill=ptabs(~gear|cyl,data=mtcars))
```

The `ptabs` function is a high level wrapper for `xtabs` and the lower level functions `marginalize` 
and `condition`.  Construct an array of joint probabilities of numbers of gears, cylinders and 
carburetors,
```{r}
P <- ptabs(~gear+carb+cyl,data=mtcars)
```
The `condition` function conditions the array on a specified set of variables.  To calculate
the conditional probability of the number of cylinders and carburetors given the number of gears
```{r}
P <- condition(P,c("gear"))
P
```
The `marginalize` function marginalizes an array to a specified set of variables.  To calculate
the probability of the number of cylinders conditional on the number of gears
```{r}
marginalize(P,c("cyl"))
```




